---
title: Introducing the partition package
author: Malcolm Barrett
date: '2019-05-20'
slug: introducing-the-partition-package
categories:
  - r
  - releases
tags: [partition]
subtitle: ''
summary: ''
authors: []
lastmod: '2019-05-20'
featured: no
image:
  caption: 'Photo by [Braden Collum](https://unsplash.com/photos/LDh4LAV2Duc)'
  focal_point: ''
  preview_only: yes
projects: [partition]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dpi = 320)
```


I'm pleased to announce the CRAN release of partition 0.1.0. partition is a fast and flexible data reduction framework that minimizes information loss and creates interpretable clusters. partition uses agglomorative clustering: it starts from the ground up, matching pairs of variables and assessing the amount of information that would be explained by their reduction. If the information is above this user-specified threshold, the data is reduced. This type of reduction is particularly useful in very redundant data, such as high-resolution genetic data.

# Creating partitions

`partition()` takes a data frame and reduces it as much as possible without creating clusters below the minimum amount of information specified in the `threshold` argument.

```{r}
# install.packages("partition")
library(partition)
set.seed(1234)
#  simulate correlated data
df <- simulate_block_data(c(3, 4, 5), lower_corr = .4, upper_corr = .6, n = 100)

#  don't accept reductions where information < .6
prt <- partition(df, threshold = .6)
prt
```

The `partition` object returned by `partition()` contains the reduced data.

```{r}
# return reduced data
partition_scores(prt)
```

You can also access the mappings of the original data to the clusters. Variables map to one and only one cluster.

```{r}
# access mapping keys
mapping_key(prt)

unnest_mappings(prt)
```

# Using partitioners

partition uses an approach called Direct-Measure-Reduce to agglomerate the data: functions called partitioners tell the algorithm 1) where to look in the data 2) how to measure information loss and 3) how to reduce the data. The default partitioner that `partition()` uses called `part_icc()`, which 1) finds the closest pair of variables using a correlation-based distance matrix 2) measures information using intraclass correlation and 3) reduces acceptable clusters using scaled rowmeans.

partition also has a number of other options. `part_kmeans()`, for instance, uses the K-means algorithm to find potential reductions and reduces to the minimum level of `k` that is still above the information threshold, measured by ICC.

```{r}
# use a lower threshold of information loss
partition(df, threshold = .5, partitioner = part_kmeans())
```

See the [introductory vignette](https://uscbiostats.github.io/partition/articles/introduction-to-partition.html) for more information on the built-in partitioners.

`partition()` is actually agnostic to the Direct-Measure-Reduce approach used. This makes partition extremely flexible. You can edit existing partitioners or create completely new ones. For instance, if we want `part_icc()` to return raw row means rather than scaled row means, we can replace the `reduce` component of the function (here, with `rowmeans()`).

```{r}
# use a custom partitioner
part_icc_rowmeans <- replace_partitioner(
  part_icc, 
  reduce = as_reducer(rowMeans)
)
partition(df, threshold = .6, partitioner = part_icc_rowmeans) 
```

`partition()` works seamlessly with changes to the partitioner. See the [vignette on extending partition](https://uscbiostats.github.io/partition/articles/extending-partition.html) for more information on customizing partitioners.

partition also supports a number of ways to visualize partitions and permutation tests; these functions all start with `plot_*()`. These functions all return ggplots and can thus be extended using ggplot2.

```{r, dpi = 320}
plot_stacked_area_clusters(df) +
  ggplot2::theme_minimal(14)
```

Install partition now from CRAN or install the [development version on GitHub](https://github.com/USCbiostats/partition).

# Learn More
* [Vignette: Introduction to partition](https://uscbiostats.github.io/partition/articles/introduction-to-partition.html)
* [Vignette: Extending partition](https://uscbiostats.github.io/partition/articles/extending-partition.html)
* [partition web site](https://uscbiostats.github.io/partition/)
* [GitHub Repo](https://github.com/USCbiostats/partition)