---
title: Data science as an atomic habit
author: Malcolm Barrett
date: '2021-01-04'
slug: data-science-as-an-atomic-habit
draft: true
categories:
  - learning
  - habits
  - r
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2021-01-04T08:38:52-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Several years ago, I lived at the Zen Center of New York City, a Zen temple focused on supporting lay practitioners--folks who practice but also have jobs, families, and so on. Living at a Zen center is inherently intensive. You follow the temple schedule--morning and evening meditation, work and meals together, and regular meditation retreats--as well as managing your other responsibilities. Yet, there's a certain ease to practicing meditation in such a place: you don't have to give it a second thought. Everything about places like ZCNYC is set up to support your practice. Your schedule is plainly spelled out for you. If you sleep in (🙋), someone will come get you up so you can join in the meditation hall (😅). Your life *systematically* supports your practice.

When I moved to California for my Ph.D., I had a hard time with my practice. I didn't lack interest or motivation: I knew from direct experience that it was important. But now *I* was responsible for making it happen. After a number of months of struggling with sitting[^sitting] every day, I decided enough was enough. I made a vow: I was going to sit every day for the rest of my life. I didn't make a commitment to a certain amount of hours or any other requirement. I just focused on consistency. 

Making a vow to sit every day completely changed my experience. Why? The answer was clear to me almost immediately: the debate in my head was gone. There was no more "should I sit?", no more "how long?" or "when?". I didn't worry about it, because it was a solved problem: I sit every day, first thing. Nothing to think about. Even though spending *more* time practicing was important to me, I didn't focus on that. I focused on the system that got me on the cushion. Because it's important to me, spending more time doing it bloomed naturally from a consistent practice; if I just got on the cushion every day, that was enough. 

I recently read [*Atomic Habits* by James Clear](https://jamesclear.com/atomic-habits), which I quite liked ([here's a summary](https://kislayverma.com/books/book-review-atomic-habits/)). A lot of books on productivity are one or two good blog posts wrapped in 200 pages of gimmicky marketing, but I found this book readable and practical, supported by research. When I read the book, I realized: oh, what I did with my practice was an atomic habit. I focused on small, consistent habits that build on their own. I ingrained them in my life so that I did not need to think about it. I *became* a person who gets on the cushion every day. 

# Data science as an Atomic Habit

In Clear's view, the most important habits are small and consistent, whose benefits slowly build and compound over time, much like a retirement account. Relatedly, we *think* our abilities should grow linearly to our effort, but that is not so. This was exactly my experience learning R and other tools for data science. When you're learning R or python, for instance, you may find yourself wondering why, if you're coding or studying every day, you still feel like the language doesn't *get* you. Clear uses a diagram to show this, but since this is a data science post, we might as well draw it in R:

```{r}
library(ggplot2)
inner_thoughts <- tibble::tribble(
  ~x,  ~y,    ~label,
  .5,   800,  "Cool!",
  1.25, 800,  "...?",
  3,    800,  "What the heck is a closure?",
  4.75, 900,  "Ugghhhh",
  6,    1200, "Ugghhhhhhh",
  7.25, 2500, "Oh wait",
  8,    4000, "...!!",
  8.75, 8000, "!!!!!",
  9.2, 20000, "Oh sure, it's easy"
)

ggplot() + 
  geom_function(fun = exp, size = 1.2, color = "firebrick") + 
  geom_text(data = inner_thoughts, aes(x, y, label = label)) +
  xlim(0, 10) + 
  ggtitle("Mastering a programming language") + 
  theme_void(24)
```

This is the stubbornly quiet path of practice: our abilities lag behind our effort. It's right at "Ugghhhh" stage that our internal motivation to keep on starts to dissolve. Sure, you can make a plot or whatever, but you're still fighting errors all day. What do you have to show for it at this point? Not that much more than when you started. But our growth is rarely linear: often, it's exponential, at least at key moments. We suddenly start to make connections, which help us build other connections, and so on. Growth is never infinite (bad news: you're going to die some day even if you improve every day until then), but it's also rarely as linked to our *immediate* efforts as we expect. 

# Practice and becoming are one action

With a really meaningful habit--like setting out to learn data science--part of the issue is that we're focused on the results, as if our (not) being a data scientist is equal to how many errors our code makes. But that's not quite right. If you're learning to run, for instance, you aren't becoming a person who runs for 30 minutes. You're becoming a *runner*. That transformation is deeper, more challenging, and richer. Running 30 minutes is just a useful strategy to engage with that transformation. When you're sitting down with code and data, you're not becoming a person who types R code without errors[^errors]. You're *becoming* a data scientist. Becoming--birth--is a slow, sometimes painful, but ultimately astounding process. In Zen, we say, "don't put a head on top of your own." Don't confuse the results with what's really happening. 

One of the better Hadley Wickham gifs is this one, [where he's warning R learners](http://datascience.la/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/) that, well, you're going to suck at R for a while:

```{r echo=FALSE}
blogdown::shortcode('tweet', '1022233480497954817')
```

The fuller quote is: 

> Whenever you’re learning a new tool, for a long time you’re going to suck... But the good news is that is typical; that’s something that happens to everyone, and it’s only temporary.

I've always liked this because you *should* prepare yourself for not being a master programmer for a while; it really is an inherent part of the process and largely unavoidable. I also think, however, that it's good to recognize that your learning, your effort--they're not going to waste. *Becoming* a data scientist merely lags behind the effort we put in to *practicing* data science. Practice and becoming are one process, one action, but we often don't see how they're connected. What's happening underneath the surface of our effort is quiet--right until it rises with a fury.

# One foot after the other

Pick something small that you can squeeze in on your busiest of days, and know when you're going to do it. This last part is important: if you don't have a set period to engage with your habit, you don't know you missed it until the end of the day, when perhaps it's too late. A useful strategy is to stack the new habit on an existing habit. Do you already make a cup of coffee every morning? Make a clear commitment attached to that habit: "I'm going to _______ every morning right after I make a cup of coffee." The habit you stack it on top of can be anything as long is it's something you do every day.

So what would an atomic habit for becoming a data scientist look like? Here are a few ideas:

1. Open RStudio and type "Good morning, R" into the console. That's it. Seriously.
2. Write code for 5 minutes
3. Use one new function. Reading the help page counts.
4. Make a single commit on git. 

These extremely small habits are useful because you can always fit them in, but they work best when you have a sense of what you might do after. When I used to run (prior to a hamstring injury), I followed a running schedule, but when I was having trouble getting off the couch, I would say to myself: "all you have to do is put on your running shoes and step out on the porch, then you can be done." Usually that was enough for me to engage with whatever my original plan was. It's also helpful if you're plan involves cool or urgent things to keep you motivated. Luckily, in the world of data science, there's no shortage of either. In other words, focus on high-impact activity. Many data scientists will benefit most from starting with data visualization and project workflow tools, while a software engineer might benefit from starting with language fundamentals.[^start]

In other words, the habit you're building is showing up. Action drives motivation. When you wait around for motivation to drive action, well, you end up waiting around a lot.

# Tracking

You should track your habit to make sure you *are* showing up. One nice feature of GitHub is that it shows your commit history in a heatmap; if you commit, it's green, and if not, it's gray. If your habit is to make a commit every day, don't leave any gray squares. This is very similar to a popular productivity technique: [The Seinfeld Chain](https://lifehacker.com/jerry-seinfelds-productivity-secret-281626). Jerry Seinfeld made a commitment to write a joke every day. He had a big, daily calendar, and he crossed off a day in red marker every time he wrote. You build up a chain of red Xs. Quoth Jerry: "Don't break the chain."

There are several apps that are good for this. I use [HabitHub](https://www.thehabithub.com/). I used to draw out 365 squares in my gridded Moleskine notebook and cross them off manually. 

Right now, I track 5 habits on HabitHub. I slowly added them, focusing on one at a time, until they felt secure, then I add another. I also made modifications to them over time as my needs for this support changed; some habits became so automatic that I didn't need to track them any more, while others benefited from creating a new, closely-related habit to stack on the one that had become stable. 

# Trust yourself

Ultimately, what you're cultivating here is consistency and patience with the process. We often conflate patience with grit. Patience, however, is trust, not endurance. It is, in particular, trust in yourself. That's good news, because keeping your nose the grindstone is useful, but it's not sustainable. Eventually you run out of nose. Instead, apply your grit to where it's most beneficial: consistency and trust. Get to the keyboard every day. Make that commit every day. Show up. While you're doing that, make the choice to see your victories. Cultivate trust in yourself by witnessing yourself *for who you are*: someone who is really capable of what you've set out to do. Witness the evidence of your abilities--perhaps small and simple at first. Then witness it bloom.

[^sitting]: The term "sitting" is a euphemism for doing meditation. It's especially common in Zen, where we practice [zazen](https://zmm.org/teachings-and-training/meditation-instructions/), a form of silent, seated meditation. In fact, "zazen" means "seated meditation" or "sitting Zen".
[^errors]: More good news/bad news: you're never going to stop writing code with bugs. Might as well make friends with it now!
[^start]: Where you start in a language depends on what your setting out to acheive, as well as your background. Many teachers of R and python start with things like object types and control flow. But those are *language* fundamentals. You'll probably need to know them eventually, but depending on your needs, you might be better off starting with *data science* fundamentals: reading data, visualizing data, manipulating data, and so on. Of course, this is totally dependent. Perhaps you have a computer science background and starting with control flow sounds great to you! 
